\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-13 NAR_DRAFT.tex
Edits by Ryan.

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
  \end{array}
\end{equation}
The first two terms on the right-hand side can be estimated from the mutation profile (Fig.~\ref{NAR-sigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{m,c}$.
The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is modeled as multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{m,c}\} $.
In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
As a result we need to model $\mathrm{p}(\mathrm{S})$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,

$$
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
$$
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a-priori a mutation is equally likely in any context.

In order to estimate $\boldsymbol{\pi}$, we generate a subset of variants in the tumor in which we have high confidence that the variants are present, based on an initial calculation of their likelihood give the data.
These are variants such that for any reasonable value of the site-specific prior probability of mutation, the evidence for the variant in the read data erases any doubt about the presence of the variant. 
Let $\mathrm{D}$ be the set of high confidence variant calls -- which we define as having posterior odds greater than 10 to 1 -- and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
  $$
Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.
The R implementation for \batcave ships with pre-computed tables for both whole exomes and whole genomes.

>>>>

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
  \end{array}
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Fig.~\ref{NAR-sigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

\rngcomment{Got here...}

We model the observed mutation profile $S$ (Fig.~\ref{NAR-sigfig}a-c) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model $\mathrm{p}(\mathrm{S})$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
$$
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
$$
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.

To estimate $\boldsymbol{\pi}$, we identify a subset of high confidence variants, based on an initial calculation of their likelihood given the data.
These are variants for which the evidence in the read data overwhelms any reasonable value of the site-specific prior probability of mutation.
Let $\mathrm{D}$ be the set of high confidence variant calls, which we define as those having posterior odds greater than 10 to 1 with the site-specific prior, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
  $$
Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  \mathrm{p}(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   \mathrm{p}(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.
The R implementation for \batcave ships with pre-computed tables for both whole exomes and whole genomes.
